Robust co-occurrence quantification for low-dimensional lexical distributional semantics

### 2016-02-25 paper.tex
MP edits

\title{Robust co-occurrence quantification for lexical distributional semantics experiments}

@@

\newcite{TACL570} suggested to optimize PMI with a  set of parameters adopted from predictive models of \newcite{mikolov2013efficient}, namely shifting and context distribution smoothing. Their experiments and thus the parameter selection recommendations use highly dimensional vector spaces with word vector  dimensionality of almost \textbf{200K}. As such, recent work on lexical distributional semantics with the results close to the state of the art involve vectors with a considerable dimensionality of hundreds of thousands \cite{baroni-dinu-kruszewski:2014:P14-1,kiela-clark:2014:CVSC,lapesa2014large}.

Contrary to the trend of employing highly dimensional vectors, other work on lexical and compositional distributional semantics employ vectors with much less dimensions, for example, \textbf{2K} \cite{Grefenstette:2011:ESC:2145432.2145580,kartsadrqpl2014,milajevs-EtAl:2014:EMNLP2014}, \textbf{3K} \cite{Dinu:2010:MDS:1870658.1870771,milajevs-purver:2014:CVSC} or \textbf{10K} \cite{polajnar-clark:2014:EACL,Baroni2010nouns}.

Such a mismatch in vector dimensionality selection across lexical and compositional tasks rises a number of questions this paper aims to answer.
\begin{itemize}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence performance of highly dimensional models the same way as low dimensional models? Can the findings of \newcite{TACL570} be directly applied to low dimensional models?
\item Is it possible to come up with the heuristics for model parameter selection? If so, what are they?
\end{itemize}

To answer these questions we perform a systematic study of distributional models with a rich set of parameters on two lexical datasets: SimLex-999 \cite{hill2014simlex} and MEN \cite{Bruni:2014:MDS:2655713.2655714}.

@@

Co-occurrence weightings schemes based on \emph{Point-wise mutual information} (PMI, Equation~\ref{eq:pmi}) have been successfully applied in distributional semantics (see for example \newcite{J90-1003}, \newcite{Turney:2010:FMV:1861751.1861756} and \newcite{NIPS2014_5477}).
%
\begin{equation}
  \label{eq:pmi}
  \operatorname{PMI}(x, y) = \log\frac{P(x,y)}{P(x)P(y)}
\end{equation}

@@

It has been experimentally shown that negative PMI values do not positively contribute to model performance and worth nullifying \cite{Turney:2010:FMV:1861751.1861756}. Positive PMI extended with an additional additional hyperparameter $k$ (\texttt{neg}) following \newcite{TACL570} is the third PMI variant (abbreviated as SPMI) used in this work:
%
\begin{equation}
  \label{eq:ppmi}
  \operatorname{SPMI_k} = \max (0, \operatorname{PMI}(x, y) - \log k)
\end{equation}
%
We can apply the same idea to CPMI:
%
\begin{equation}
  \label{eq:pcpmi}
  \operatorname{SCPMI_k} = \max (0, \operatorname{CPMI}(x, y) - \log 2k)
\end{equation}

@@

We test a widely used cosine based similarity, as well as correlation \cite{kiela-clark:2014:CVSC}.

@@

\subsection{Heurisitc evaluation}
\label{sec:heurisitc-evaluation}

@@

This paper presents a systematic study of co-occurrence quantification focusing of the selection of parameters presented in \newcite{TACL570}. We replicate their recommendation for highly dimensional vector spaces, and show that with appropriate parameter selection it is possible to achieve comparable performance with low dimensional spaces. As an outcome of such an analysis we propose a set of model selection heuristics that maximize performance.

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). It is beneficial to experiment with dense and unsmoothed low dimensional spaces ($k < 1$, global context probability); and sparse and smooth highly dimensional spaces ($k > 1$, $\alpha = 0.75$). However, these heuristics do not guarantee the best possible result for unweighted and $n$-weighted models because of the high variance of the corresponding scores. Based on this we suggest to use \logNSPMI/ or \logNSCPMI/ with the dimensionality of at least 20K to ensure good performance on lexical tasks.

>>>>

%\title{Robust co-occurrence quantification for lexical distributional semantics experiments}
% MP: how about:
\title{Robust co-occurrence quantification for low-dimensional lexical distributional semantics}

@@

\newcite{TACL570} propose optimisations for co-occrrence-based distributional models, using parameters adopted from predictive models \cite{mikolov2013efficient}: \emph{shifting} and \emph{context distribution smoothing}. Their experiments and thus their parameter  recommendations use high-dimensional vector spaces with word vector dimensionality of almost \textbf{200K}, and many recent state-of-the-art results in lexical distributional semantics have been obtained using vectors with similar dimensionality  \cite{baroni-dinu-kruszewski:2014:P14-1,kiela-clark:2014:CVSC,lapesa2014large}.

In contrast, much work on \emph{compositional} distributional semantics employs vectors with much fewer dimensions: e.g.~\textbf{2K} \cite{Grefenstette:2011:ESC:2145432.2145580,kartsadrqpl2014,milajevs-EtAl:2014:EMNLP2014}, \textbf{3K} \cite{Dinu:2010:MDS:1870658.1870771,milajevs-purver:2014:CVSC} or \textbf{10K} \cite{polajnar-clark:2014:EACL,Baroni2010nouns}. \textbf{MP: say why}

Such a mismatch in vector dimensionality between lexical and compositional tasks gives rise to a number of questions:
\begin{itemize}
\item To what extent does model performance depend on vector dimensionality?
\item Do parameters influence performance of high-dimensional and low-dimensional models in the same way? Can the findings of \newcite{TACL570} be directly applied to low-dimensional models?
\item If not, can we derive suitable parameter selection heuristics which take account of dimensionality?
\end{itemize}

To answer these, we perform a systematic study of distributional models with a rich set of parameters on two lexical datasets: SimLex-999 \cite{hill2014simlex} and MEN \cite{Bruni:2014:MDS:2655713.2655714}.

@@

Co-occurrence weighting schemes based on \emph{Point-wise mutual information} (PMI, Equation~\ref{eq:pmi}) have been successfully applied in distributional semantics (see e.g.~\newcite{J90-1003}, \newcite{Turney:2010:FMV:1861751.1861756}, \newcite{NIPS2014_5477}).
%
\begin{equation}
  \label{eq:pmi}
  \operatorname{PMI}(x, y) = \log\frac{P(x,y)}{P(x)P(y)}
\end{equation}

@@

Many approaches use only \emph{positive} PMI values, as  negative PMI values may not positively contribute to model performance \cite{Turney:2010:FMV:1861751.1861756}. This can be generalised as an additional hyperparameter $k$ (\texttt{neg}) following \newcite{TACL570}, giving our third PMI variant (abbreviated as SPMI):
%
\begin{equation}
  \label{eq:ppmi}
  \operatorname{SPMI_k} = \max (0, \operatorname{PMI}(x, y) - \log k)
\end{equation}
%
We can apply the same idea to CPMI:
%
\begin{equation}
  \label{eq:pcpmi}
  \operatorname{SCPMI_k} = \max (0, \operatorname{CPMI}(x, y) - \log 2k)
\end{equation}

@@

We test a widely used cosine based similarity, as well as correlation \cite{kiela-clark:2014:CVSC}. \textbf{MP: I would move this to next section: it's not really a parameter here}

@@

\subsection{Evaluation of heuristics}
\label{sec:heuristic-evaluation}

@@

This paper presents a systematic study of co-occurrence quantification focusing of the selection of parameters presented in \newcite{TACL570}. We replicate their recommendation for high-dimensional vector spaces, and show that with appropriate parameter selection it is possible to achieve comparable performance with low-dimensional spaces. As an outcome of such an analysis we propose a set of model selection heuristics that maximize performance.

In general, model performance depends on vector dimensionality (the best setup with 50K dimensions is better than the best setup with 1K dimensions by 0.03 on SimLex-999). It is beneficial to experiment with dense and unsmoothed low-dimensional spaces ($k < 1$, global context probability); and sparse and smooth high-dimensional spaces ($k > 1$, $\alpha = 0.75$). However, these heuristics do not guarantee the best possible result for unweighted and $n$-weighted models because of the high variance of the corresponding scores. Based on this we suggest to use \logNSPMI/ or \logNSCPMI/ with the dimensionality of at least 20K to ensure good performance on lexical tasks.
